Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-23 paper/paper.tex
fix some type theory stuff, non-linear types intro, cite vakar, missing \otimes term syntax fixed

\begin{figure}
\begin{mathpar}
  \inferrule
  {
    \inferrule
    {~}
    {\cdot \vdash \nil : \A^*}
    \\
    \inferrule
    {
      \inferrule
      {
        \inferrule
        {~}
        {a_1 : \A, a_2 : \A , a_{*} : \A^* \vdash \cons~a_1 (\cons~a_2~a_{*}) : \A^*}
      }
      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {a_1 \otimes a_2}
        {a'} {\cons~a_1 (\cons~a_2~a_{*}) : \A^*}}
    }
    {\cdot \vdash f := \lamblto a {\lamblto {a_{*}} {\app {\app g a} {a_{*}} }} : (\A \otimes \A) \lto \A^* \lto \A^*}
  }
  {a : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(a) : \A^* }
\end{mathpar}
\begin{lstlisting}
h : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
h nil = nil
h (cons (a1 , a2) as) = cons a1 (cons a2 (h as))
\end{lstlisting}
\caption{A parse transformer for abstract grammars}
\label{fig:kleeneabstractproof}
\end{figure}

@@

\steven{TODO define general inductives as the least fixed point of strictly
  positive endofunctors on $U$}

\steven{TODO linear $\mu$ in the type formation rules? nonlinear $\mu$ in the
  type formation rules?}

\begin{figure}
  \begin{mathpar}
    \footnotesize
    \boxed{\ctxwffjdg \Gamma X}

@@

    % \inferrule{~}{\ctxwffjdg \Gamma {Nat}}

@@

    \boxed{\ctxwffjdg \Gamma {X \equiv Y}}

@@

    \boxed{\Gamma \vdash X \isSmall}

    \inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash X \isSmall}

    \boxed{\linctxwffjdg \Gamma A}

@@

    \boxed{\linctxwffjdg \Gamma {A \equiv B}}

@@

    \inferrule{\linctxwffjdg \Gamma A}{\linctxwffjdg \Gamma {\unquoteTy
        {\quoteTy A} \equiv A}}
  \end{mathpar}
  \caption{Type formation rules}
  \label{fig:formation}
\end{figure}

\begin{figure}
  \begin{mathpar}
  \footnotesize
    \boxed{\nonlinterm \Gamma M X}

    \inferrule{\nonlinterm \Gamma {M \equiv N} {X}}
      {\nonlinterm \Gamma {refl} {M =_{X} N}}
  \end{mathpar}
  \caption{Non-linear typing (selection)}
  \label{fig:non-linear-types}
\end{figure}

\subsection{Non-linear Typing}
The non-linear types of \theoryabbv include the standard dependent
types for $\Pi$,$\Sigma$, extensional equality, unit and empty types, and
inductive types that are defined strictly positively. We may also assume function
extensionality\cite{Hofmann_1997}; Function
extensionality is holds in Cubical Agda, and in our implementation we make heavy use of it of when
validating that the denotational semantics respects the equational theory of
\theoryabbv. However, it is not necessary to assume function extensionality for
the non-linear types in \theoryabbv, as we don't utilize it in any of our constructions.

We present a selection of
the non-linear type constructions in
\Cref{fig:non-linear-types} and provide the rest in
\cref{fig:full-non-linear-types, fig:jdg-eq-nonlinear} of the appendix.
First, we include universe types $U$ of
small non-linear types and $L$ of linear types. These are defined as
universes ``ala Coquand'' in that we define a judgment saying when a
non-linear type is small and define the universe to internalize
precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. These universe
types are needed so that we can define types by recursion on natural
numbers. As in \cite{krishnaswami_integrating_2015}, the extensional equality type
$M =_{X} N$ used here has the introduction form $refl$, but no
elimination due to the presence of equality reflection. Next, we include a non-linear type $\ltonl A$ where $A$ is a
linear type. The intuition for this type is that its elements are the
linear terms that are ``resource free'': its introduction rule says we
can construct an $\ltonl A$ when we have a linear term of type $A$
with no free linear variables. This type is used extensively in our
examples, playing a similar role to the $!$ modality of ordinary
linear logic or the persistence modality $\square$ of separation logic
\cite{girard_linear_1987,jung_higher-order_2016}.

\subsection{Linear Typing}
We give the rules for linear type
formation in \cref{fig:formation} and the definition of linear terms in
\cref{fig:linear-terms}. The equational theory for these
types is straightforward $\beta\eta$ equivalence and included in the
appendix. First, the linear variable rule says that a linear variable
can be used if it is the \emph{only} variable in the context.

@@

\steven{I'm not sure if this is the proper way to cite the corollary, especially
because the citation is a reference, not the originator of the idea}
\begin{corollary}[\cite{Cockett_1993}]
  \label{cor:binary-mono}
  Distributivity implies that the coprojections $\inl : A \to A \oplus B$,
  $\inr : B \to A \oplus B$ of a binary sum are monomorphisms --- i.e. if
  $\inl\,a \equiv \inl\,a'$, then $a \equiv a'$.
\end{corollary}

@@

Our syntax for non-commutative dependent linear type theory is based on the
dependent commutative linear type theory of
\citet{krishnaswami_integrating_2015}, itself an extension of Benton and
Wadler's linear-non-linear calculus \cite{bentonMixedLinearNonlinear1995}. A
distinct feature of these systems is that linear types and non-linear types are
distinct sorts, so the linear logic $!$ operator is not a primitive, but is
instead definable. Vakar develops a dependent linear type theory where instead
the non-linear types are accessed using the $!$ modality, similar to Girard's
original approach \cite{vakar,girard_linear_1987}. Additionally, Vakar develops
a general categorical semantics for their system, whereas we have only developed
a single denotational model. It appears straightforward to adapt Vakar's general
semantics approach to a non-commutative variant that would apply to
\theoryabbv. This may have applications in finding alternative models, or
developing logical relations proofs categorically.

@@

\begin{figure}
  \begin{mathpar}
  \footnotesize
    \boxed{\nonlinterm \Gamma M X}

@@

    \inferrule{\linterm \Gamma \cdot e A}{\nonlinterm \Gamma e \ltonl{A}}
  \end{mathpar}
  \caption{Non-linear Typing}
  \label{fig:full-non-linear-types}
\end{figure}

>>>>

\begin{figure}
\begin{mathpar}
  \inferrule
  {
    \inferrule
    {~}
    {\cdot \vdash \nil : \A^*}
    \\
    \inferrule
    {
      \inferrule
      {
        \inferrule
        {~}
        {a_1 : \A, a_2 : \A , a_{*} : \A^* \vdash \cons~a_1 (\cons~a_2~a_{*}) : \A^*}
      }
      {a' : \A \otimes \A , a_{*} : \A^* \vdash g := \letin {(a_1 , a_2)}
        {a'} {\cons~a_1 (\cons~a_2~a_{*}) : \A^*}}
    }
    {\cdot \vdash f := \lamblto a {\lamblto {a_{*}} {\app {\app g a} {a_{*}} }} : (\A \otimes \A) \lto \A^* \lto \A^*}
  }
  {a : (\A \otimes \A)^* \vdash \texttt{h} := \fold(\nil , f)(a) : \A^* }
\end{mathpar}
\begin{lstlisting}
h : (*@$\uparrow$@*)((A (*@$\otimes$@*) A)(*@$^*$@*) (*@$\lto$@*) A(*@$^*$@*))
h nil = nil
h (cons (a1 , a2) as) = cons a1 (cons a2 (h as))
\end{lstlisting}
\caption{A parse transformer for abstract grammars}
\label{fig:kleeneabstractproof}
\end{figure}

@@

\subsection{Non-linear Typing}

\begin{figure}
  \begin{mathpar}
  \footnotesize
    \boxed{\inferrule{\ctxwff\Gamma}{\ctxwffjdg \Gamma X}}

@@

    \inferrule{~}{\ctxwffjdg \Gamma {Nat}}

@@

    \boxed{\inferrule{\ctxwffjdg\Gamma X \and \ctxwffjdg\Gamma Y}{\ctxwffjdg \Gamma {X \equiv Y}}}

@@

    \boxed{\inferrule{\ctxwffjdg \Gamma X}{\Gamma \vdash X \isSmall}}

    \\
    \boxed{\ctxwffjdg \Gamma X}{\nonlinterm \Gamma M X}

    \inferrule{\nonlinterm \Gamma {M \equiv N} {X}}
      {\nonlinterm \Gamma {refl} {M =_{X} N}}

    \\
    \boxed{\inferrule{\nonlinterm\Gamma M X \and \nonlinterm \Gamma N X}{\nonlinterm \Gamma {M \equiv N} X}}

    \inferrule{\nonlinterm \Gamma P {M =_{X} N}}{\nonlinterm\Gamma{M \equiv N} X}
  \end{mathpar}
  \caption{Non-linear Formation and Typing Rules (selection)}
  \label{fig:non-linear-types}
\end{figure}

We present a selection of the non-linear type constructions in
\Cref{fig:non-linear-types} and provide the rest in
\cref{fig:full-non-linear-types, fig:jdg-eq-nonlinear} of the appendix.  First,
we include universe types $U$ of small non-linear types and $L$ of linear
types. These are defined as
universes ``ala Coquand'' in that we define a judgment saying when a
non-linear type is small and define the universe to internalize
precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. These universe
types are needed so that we can define types by recursion on natural
numbers. 
Next, we include standard $\Sigma,\Pi$, empty, unit and natural number types.

We use an \emph{extensional} equality type $M =_{X} N$ with introduction form
$refl$, but no elimination form. Instead we have the equality reflection rule
which allows us to conclude a definitional equality $M =_{X} N$ from an
arbitrary typal equality proof $P$.  The usage of an extensional equality type
matches our implementation, which interprets both judgmental and typal equality
as Cubical Agda's \texttt{Path} type, and so naturally supports the equality
reflection rule. Extensional equality makes type checking of our syntax as such
undecidable because the conversion rule may require an arbitrarily complex
equality proof with no explicit proof term, but in our Agda implementation we
must provide all of these equalities manually, and so the extensionality does
not raise any issues \cite{Hofmann_1997}.

Lastly, we include a non-linear type $\ltonl A$ where $A$ is a linear type. The
intuition for this type is that its elements are the linear terms that are
``resource free'': its introduction rule says we can construct an $\ltonl A$
when we have a linear term of type $A$ with no free linear
variables. Semantically, this is the type of parses of the empty string. This
type is used extensively in our examples, playing a similar role to the $!$
modality of ordinary linear logic or the persistence modality $\square$ of
separation logic \cite{girard_linear_1987,jung_higher-order_2016}.

\subsection{Linear Typing}

\begin{figure}
  \begin{mathpar}
    \footnotesize
    \boxed{\inferrule{\ctxwff\Gamma}{\linctxwffjdg \Gamma A}}

@@

    \\
    \boxed{\inferrule{\linctxwffjdg \Gamma A \and \linctxwffjdg \Gamma B}{\linctxwffjdg \Gamma {A \equiv B}}}

@@

    \inferrule{\linctxwffjdg \Gamma A}{\linctxwffjdg \Gamma {\unquoteTy
        {\quoteTy A} \equiv A}}
  \end{mathpar}
  \caption{Linear Type Formers, Type Equivalence}
  \label{fig:linear-formation}
\end{figure}

We give the rules for linear type
formation in \cref{fig:linear-formation} and the definition of linear terms in
\cref{fig:linear-terms}. The equational theory for these
types is straightforward $\beta\eta$ equivalence and included in the
appendix. First, the linear variable rule says that a linear variable
can be used if it is the \emph{only} variable in the context.

@@

The following corollary is a well known consequence of distributivity \cite{Cockett_1993}.
\begin{corollary}
  \label{cor:binary-mono}
  Distributivity implies that the constructors $\inl : A \to A \oplus B$,
  $\inr : B \to A \oplus B$ of a binary sum are injective --- i.e. if
  $\inl\,a \equiv \inl\,a'$, then $a \equiv a'$.
\end{corollary}

@@

Our syntax for non-commutative dependent linear type theory is based on the
dependent commutative linear type theory of
\citet{krishnaswami_integrating_2015}, itself an extension of Benton and
Wadler's linear-non-linear calculus \cite{bentonMixedLinearNonlinear1995}. A
distinct feature of these systems is that linear types and non-linear types are
distinct sorts, so the linear logic $!$ operator is not a primitive, but is
instead definable, for instance in \theoryabbv as $!A = \bigoplus_{\_:\ltonl A}
I$. Vakar develops a dependent linear type theory where instead the non-linear
types are accessed using the $!$ modality, similar to Girard's original approach
\cite{vakar,girard_linear_1987}. Additionally, Vakar develops a general
categorical semantics for their system, whereas we have only developed a single
denotational model. It appears straightforward to adapt Vakar's general
semantics approach to a non-commutative variant that would apply to
\theoryabbv. This may have applications in finding alternative models, or
developing logical relations proofs categorically.

@@

\max{TODO: include the rest of the rules}
\begin{figure}
  \begin{mathpar}
  \footnotesize
    \boxed{\nonlinterm \Gamma M X}

@@

  \end{mathpar}
  \caption{Non-linear Typing}
  \label{fig:full-non-linear-types}
\end{figure}
